---
title: "DAPC on Himalayan Birds data"
author: "Kushal K Dey"
date: "February 29, 2016"
output: 
  html_document:
    css: floating-toc.css
    toc: true
---

## Overview

In this script, we perform DAPC (Dicriminant analysis of Principal Components) on the Himalayan birds data. 

```{r echo=TRUE, eval=TRUE}

data=read.csv("../external_data/Himalayan_grid_matrix.csv",header=TRUE);

counts=as.matrix(data[,-1]);

rownames(counts)=data[,1];

new_data1 <- data.frame(read.csv('../external_data/MohanGrids.csv'));
new_data2 <- data.frame(read.csv('../external_data/MohanGrids2.csv'));

bird_species <- union(as.vector(colnames(counts)), union(as.vector(new_data1[,1]), as.vector(new_data2[,1])));

new_data <- matrix(0,dim(counts)[1]+3,length(bird_species));
new_data[1,match(new_data1[,1],bird_species)]=new_data1[,2];
new_data[2,match(new_data1[,1],bird_species)]=new_data1[,3];
new_data[3,match(new_data2[,1],bird_species)]=new_data2[,2];
new_data[4:(dim(counts)[1]+3),match(colnames(counts),bird_species)]=counts;

new_counts <- as.matrix(new_data);

rownames(new_counts) <- c(c("U1","U2","MA1"),rownames(counts));
colnames(new_counts) <- bird_species;
new_counts <- new_counts[-(1:3),];

```

## Loading metadata

The metadata rows are arranged in the same order as in the data (checked).

```{r echo=TRUE, eval=TRUE}
metadata=read.csv("../external_data/Himalayan_grid_metadata.csv",header=TRUE);
elevation_metadata=metadata$Elevation;

east_west_dir = metadata$WorE;

```


## 2 cluster model

```{r echo=TRUE, eval=TRUE}
library("adegenet")
n.clust <- 2
grp <- find.clusters(new_counts, max.n=35, n.pca=200, scale=FALSE,
n.clust=n.clust) # takes about 2 minutes
dapc_out <- dapc(data.frame(new_counts), grp$grp, n.pca=10, n.da=7, center=TRUE,
     scale=FALSE, var.contrib=TRUE, var.loadings=FALSE, pca.info=TRUE,
     pca.select=c("nbEig","percVar"), perc.pca=NULL)


myCol_list <- c("lightblue","green","purple","orange", "red", "blue")
myCol <- myCol_list[1:n.clust];
scatter(dapc_out, posi.da="bottomright", bg="white", col=myCol, pch=17:22)
dapc_out$posterior

compoplot(dapc_out, posi="bottomright",
txt.leg=paste("Cluster", 1:4), lab="",
ncol=2, xlab="individuals", col=myCol)

```


### Ordering dapc clusters by elevation

```{r echo=TRUE, eval=TRUE}
K <- 4
barplot(t(dapc_out$posterior[order(elevation_metadata),]),col=myCol,axisnames=F,space=0,border=NA,main="",las=1,ylim=c(0,1),cex.axis=1.5,cex.main=1.4, xlab="elevation")
title(main=paste("Structure Plot topic proportions,k=",K))
combo_patch_dir = paste0(round(elevation_metadata));
combo_patch_dir_ordered = combo_patch_dir[order(elevation_metadata)];

match_labs=match(unique(combo_patch_dir_ordered),combo_patch_dir_ordered);
match_labs_suffix=c(match_labs[2:length(unique(combo_patch_dir_ordered))],35);
match_labs_prefix=match_labs[1:(length(unique(combo_patch_dir_ordered)))];
labs=match_labs_prefix + 0.5*(match_labs_suffix - match_labs_prefix);
  
axis(1,at=labs,unique(combo_patch_dir_ordered),las=2);
abline(v=match_labs[2:length(match_labs)]);
```

### Ordering dapc clusters by WorE

```{r echo=TRUE, eval=TRUE}
east_west_elevation = paste0(metadata$WorE, "_", metadata$Elevation);

index1 <- which(metadata$WorE=="E");
index2 <- which(metadata$WorE=="W");
elevation1 <- metadata$Elevation[index1]; elevation2 <- metadata$Elevation[index2];
index_WE <- c(index1[order(elevation1)], index2[order(elevation2)]);

    barplot(t(dapc_out$posterior[index_WE,]),col=myCol,axisnames=F,space=0,border=NA,main=paste("No. of clusters=",length(myCol)),las=1,ylim=c(0,1),cex.axis=1.5,cex.main=1.4)

    combo_patch_dir = paste0(east_west_elevation);
    combo_patch_dir_ordered = combo_patch_dir[index_WE];

  match_labs=match(unique(combo_patch_dir_ordered),combo_patch_dir_ordered);
  match_labs_suffix=c(match_labs[2:length(unique(combo_patch_dir_ordered))],35);
  match_labs_prefix=match_labs[1:(length(unique(combo_patch_dir_ordered)))];
  labs=match_labs_prefix + 0.5*(match_labs_suffix - match_labs_prefix);
  
  axis(1,at=labs,unique(combo_patch_dir_ordered),las=2);
  abline(v=match_labs[2:length(match_labs)]);
  
```


## 3 cluster model

```{r echo=TRUE, eval=TRUE}
library("adegenet")
n.clust <- 3
grp <- find.clusters(new_counts, max.n=35, n.pca=200, scale=FALSE,
n.clust=n.clust) # takes about 2 minutes
dapc_out <- dapc(data.frame(new_counts), grp$grp, n.pca=10, n.da=7, center=TRUE,
     scale=FALSE, var.contrib=TRUE, var.loadings=FALSE, pca.info=TRUE,
     pca.select=c("nbEig","percVar"), perc.pca=NULL)


myCol_list <- c("lightblue","green","purple","orange", "red", "blue")
myCol <- myCol_list[1:n.clust];
scatter(dapc_out, posi.da="bottomright", bg="white", col=myCol, pch=17:22)
dapc_out$posterior

compoplot(dapc_out, posi="bottomright",
txt.leg=paste("Cluster", 1:4), lab="",
ncol=2, xlab="individuals", col=myCol)

```


### Ordering dapc clusters by elevation

```{r echo=TRUE, eval=TRUE}
K <- 4
barplot(t(dapc_out$posterior[order(elevation_metadata),]),col=myCol,axisnames=F,space=0,border=NA,main="",las=1,ylim=c(0,1),cex.axis=1.5,cex.main=1.4, xlab="elevation")
title(main=paste("Structure Plot topic proportions,k=",K))
combo_patch_dir = paste0(round(elevation_metadata));
combo_patch_dir_ordered = combo_patch_dir[order(elevation_metadata)];

match_labs=match(unique(combo_patch_dir_ordered),combo_patch_dir_ordered);
match_labs_suffix=c(match_labs[2:length(unique(combo_patch_dir_ordered))],35);
match_labs_prefix=match_labs[1:(length(unique(combo_patch_dir_ordered)))];
labs=match_labs_prefix + 0.5*(match_labs_suffix - match_labs_prefix);
  
axis(1,at=labs,unique(combo_patch_dir_ordered),las=2);
abline(v=match_labs[2:length(match_labs)]);
```

### Ordering dapc clusters by WorE

```{r echo=TRUE, eval=TRUE}
east_west_elevation = paste0(metadata$WorE, "_", metadata$Elevation);

index1 <- which(metadata$WorE=="E");
index2 <- which(metadata$WorE=="W");
elevation1 <- metadata$Elevation[index1]; elevation2 <- metadata$Elevation[index2];
index_WE <- c(index1[order(elevation1)], index2[order(elevation2)]);

    barplot(t(dapc_out$posterior[index_WE,]),col=myCol,axisnames=F,space=0,border=NA,main=paste("No. of clusters=",length(myCol)),las=1,ylim=c(0,1),cex.axis=1.5,cex.main=1.4)

    combo_patch_dir = paste0(east_west_elevation);
    combo_patch_dir_ordered = combo_patch_dir[index_WE];

  match_labs=match(unique(combo_patch_dir_ordered),combo_patch_dir_ordered);
  match_labs_suffix=c(match_labs[2:length(unique(combo_patch_dir_ordered))],35);
  match_labs_prefix=match_labs[1:(length(unique(combo_patch_dir_ordered)))];
  labs=match_labs_prefix + 0.5*(match_labs_suffix - match_labs_prefix);
  
  axis(1,at=labs,unique(combo_patch_dir_ordered),las=2);
  abline(v=match_labs[2:length(match_labs)]);
  
```

## 4 cluster model

```{r echo=TRUE, eval=TRUE}
library("adegenet")
n.clust <- 4
grp <- find.clusters(new_counts, max.n=35, n.pca=200, scale=FALSE,
n.clust=n.clust) # takes about 2 minutes
dapc_out <- dapc(data.frame(new_counts), grp$grp, n.pca=10, n.da=7, center=TRUE,
     scale=FALSE, var.contrib=TRUE, var.loadings=FALSE, pca.info=TRUE,
     pca.select=c("nbEig","percVar"), perc.pca=NULL)


myCol_list <- c("lightblue","green","purple","orange", "red", "blue")
myCol <- myCol_list[1:n.clust];
scatter(dapc_out, posi.da="bottomright", bg="white", col=myCol, pch=17:22)
dapc_out$posterior

compoplot(dapc_out, posi="bottomright",
txt.leg=paste("Cluster", 1:4), lab="",
ncol=2, xlab="individuals", col=myCol)

```


### Ordering dapc clusters by elevation

```{r echo=TRUE, eval=TRUE}
K <- 4
barplot(t(dapc_out$posterior[order(elevation_metadata),]),col=myCol,axisnames=F,space=0,border=NA,main="",las=1,ylim=c(0,1),cex.axis=1.5,cex.main=1.4, xlab="elevation")
title(main=paste("Structure Plot topic proportions,k=",K))
combo_patch_dir = paste0(round(elevation_metadata));
combo_patch_dir_ordered = combo_patch_dir[order(elevation_metadata)];

match_labs=match(unique(combo_patch_dir_ordered),combo_patch_dir_ordered);
match_labs_suffix=c(match_labs[2:length(unique(combo_patch_dir_ordered))],35);
match_labs_prefix=match_labs[1:(length(unique(combo_patch_dir_ordered)))];
labs=match_labs_prefix + 0.5*(match_labs_suffix - match_labs_prefix);
  
axis(1,at=labs,unique(combo_patch_dir_ordered),las=2);
abline(v=match_labs[2:length(match_labs)]);
```

### Ordering dapc clusters by WorE

```{r echo=TRUE, eval=TRUE}
east_west_elevation = paste0(metadata$WorE, "_", metadata$Elevation);

index1 <- which(metadata$WorE=="E");
index2 <- which(metadata$WorE=="W");
elevation1 <- metadata$Elevation[index1]; elevation2 <- metadata$Elevation[index2];
index_WE <- c(index1[order(elevation1)], index2[order(elevation2)]);

    barplot(t(dapc_out$posterior[index_WE,]),col=myCol,axisnames=F,space=0,border=NA,main=paste("No. of clusters=",length(myCol)),las=1,ylim=c(0,1),cex.axis=1.5,cex.main=1.4)

    combo_patch_dir = paste0(east_west_elevation);
    combo_patch_dir_ordered = combo_patch_dir[index_WE];

  match_labs=match(unique(combo_patch_dir_ordered),combo_patch_dir_ordered);
  match_labs_suffix=c(match_labs[2:length(unique(combo_patch_dir_ordered))],35);
  match_labs_prefix=match_labs[1:(length(unique(combo_patch_dir_ordered)))];
  labs=match_labs_prefix + 0.5*(match_labs_suffix - match_labs_prefix);
  
  axis(1,at=labs,unique(combo_patch_dir_ordered),las=2);
  abline(v=match_labs[2:length(match_labs)]);
  
```

## Conclusions

The DAPC does not seem to be very successful in detecting biologically meaningful clusters. This may be because of the count structure of the data and the sparsity considerations that DAPC does not model and topic models do.
