---
title: "Topic Model Discrepencies"
author: "Alex White"
date: "March 18, 2016"
output: html_document
---

```{r}
######## READ IN DATA #######

library(maptpx)
library(CountClust)

data=read.csv("../external_data/Himalayan_grid_matrix.csv",header=TRUE);

counts=as.matrix(data[,-1]);

rownames(counts)=data[,1];

new_data1 <- data.frame(read.csv('../external_data/MohanGrids.csv'));
new_data2 <- data.frame(read.csv('../external_data/MohanGrids2.csv'));

bird_species <- union(as.vector(colnames(counts)), union(as.vector(new_data1[,1]), as.vector(new_data2[,1])));

new_data <- matrix(0,dim(counts)[1]+3,length(bird_species));
new_data[1,match(new_data1[,1],bird_species)]=new_data1[,2];
new_data[2,match(new_data1[,1],bird_species)]=new_data1[,3];
new_data[3,match(new_data2[,1],bird_species)]=new_data2[,2];
new_data[4:(dim(counts)[1]+3),match(colnames(counts),bird_species)]=counts;

new_counts <- as.matrix(new_data);

rownames(new_counts) <- c(c("U1","U2","MA1"),rownames(counts));
colnames(new_counts) <- bird_species;
new_counts <- new_counts[-(1:3),];

metadata=read.csv("../external_data/Himalayan_grid_metadata.csv",header=TRUE,row.names = 1);
elevation_metadata=metadata$Elevation[match(rownames(new_counts),rownames(metadata))];
east_west_dir = metadata$WorE

cbbPalette <- c("#000000", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")

forest_patch_label <- rownames(metadata)
```

There will be TWO ways of procuding the data, they should have the same content, but they are producing two different omega matrices for K=4.  I will show the discrepencies below.  The first way to input the data is using the new_counts matrix from above.  I will also show that these two different ways of inputting the data produce the SAME k=3 result, meaning there must be a problem with the model when K=4, and not some problem with plotting or somthing like that.  

The omega matrices for K=3 and K=4 are shown below. 

```{r}
#Check.BF <- topics(new_counts, K=c(2,3,4,5,6), tol=0.00001, kill=0)
K.3 <- topics(new_counts, K=3, tol=0.00001, kill=0)
#K.3$omega
K.4 <- topics(new_counts, K=4, tol=0.00001, kill=0)
#K.4$omega
```

The strucutre plots for K=3 and K=4 are shown below. 

```{r}
K <- 3
barplot(t(K.3$omega[order(elevation_metadata),]),col=cbbPalette[2:(K+1)],axisnames=F,space=0,border=NA,
        main="",las=1,ylim=c(0,1),cex.axis=1.5,cex.main=1.4)
title(main=paste("Taxonomic Assemblage Motifs, K =",K))
combo_patch_dir = paste0(forest_patch_label,"_",east_west_dir);
combo_patch_dir_ordered = combo_patch_dir[order(east_west_dir)];
match_labs=match(unique(combo_patch_dir_ordered),combo_patch_dir_ordered);
match_labs_suffix=c(match_labs[2:length(unique(combo_patch_dir_ordered))],35);
match_labs_prefix=match_labs[1:(length(unique(combo_patch_dir_ordered)))];
labs=match_labs_prefix + 0.5*(match_labs_suffix - match_labs_prefix);
labs=c(0.5,1.5,2.5,3.5,4.5,5.5,6.5,7.5,8.5,9.5,10.5,11.5,12.5,13.5,14.5,15.5,16.5,17.5,18.5,19.5,20.5,21.5,22.5,23.5,24.5,25.5,26.5,27.5,28.5,29.5,30.5,31.5,32.5,33.5,34.5)
axis(1,at=labs,unique(combo_patch_dir_ordered),las=2);
abline(v=match_labs[2:length(match_labs)]);

K <- 4
barplot(t(K.4$omega[order(elevation_metadata),]),col=cbbPalette[2:(K+1)],axisnames=F,space=0,border=NA,
        main="",las=1,ylim=c(0,1),cex.axis=1.5,cex.main=1.4)
title(main=paste("Taxonomic Assemblage Motifs, K =",K))
combo_patch_dir = paste0(forest_patch_label,"_",east_west_dir);
combo_patch_dir_ordered = combo_patch_dir[order(east_west_dir)];
match_labs=match(unique(combo_patch_dir_ordered),combo_patch_dir_ordered);
match_labs_suffix=c(match_labs[2:length(unique(combo_patch_dir_ordered))],35);
match_labs_prefix=match_labs[1:(length(unique(combo_patch_dir_ordered)))];
labs=match_labs_prefix + 0.5*(match_labs_suffix - match_labs_prefix);
labs=c(0.5,1.5,2.5,3.5,4.5,5.5,6.5,7.5,8.5,9.5,10.5,11.5,12.5,13.5,14.5,15.5,16.5,17.5,18.5,19.5,20.5,21.5,22.5,23.5,24.5,25.5,26.5,27.5,28.5,29.5,30.5,31.5,32.5,33.5,34.5)
axis(1,at=labs,unique(combo_patch_dir_ordered),las=2);
abline(v=match_labs[2:length(match_labs)]);
```

Note that for K=4 the mid and low elevation west communties form their own unique motif (in brown).  But we can read in the data differently and produce a different result for K=4 with a different omega matrix.  These new results are shown below.  All we will do is reduce the columns in the data matrix by excluding species with zero column sums. These were included in previous surveys, but are not accounted for in these data and are therefore entirely absent.  We will call the new data matrix "No zero counts"

```{r}
No_zero_counts <- new_counts[,colnames(new_counts[,which(colSums(new_counts) > 0)])]
```

Run a new topic model for K = 3 and K = 4 on No zero counts. The omega matrix for K=3 anf K=4 below.  

```{r}
#Check.BF.second <- topics(No_zero_counts, K=c(2,3,4,5,6), tol=0.00001, kill=0)
new.K.3 <- topics(No_zero_counts, K=3, tol=0.00001, kill=0)
#new.K.3$omega
new.K.4 <- topics(No_zero_counts, K=4, tol=0.00001, kill=0)
#new.K.4$omega
```

The omega matrix for K=3 is similar to the original result above, but the K=4 result is clearly different.  And now the structure plot for both Ks.  

```{r}
K <- 3
barplot(t(new.K.3$omega[order(elevation_metadata),]),col=cbbPalette[2:(K+1)],axisnames=F,space=0,border=NA,
        main="",las=1,ylim=c(0,1),cex.axis=1.5,cex.main=1.4)
title(main=paste("Taxonomic Assemblage Motifs, K =",K))
combo_patch_dir = paste0(forest_patch_label,"_",east_west_dir);
combo_patch_dir_ordered = combo_patch_dir[order(east_west_dir)];
match_labs=match(unique(combo_patch_dir_ordered),combo_patch_dir_ordered);
match_labs_suffix=c(match_labs[2:length(unique(combo_patch_dir_ordered))],35);
match_labs_prefix=match_labs[1:(length(unique(combo_patch_dir_ordered)))];
labs=match_labs_prefix + 0.5*(match_labs_suffix - match_labs_prefix);
labs=c(0.5,1.5,2.5,3.5,4.5,5.5,6.5,7.5,8.5,9.5,10.5,11.5,12.5,13.5,14.5,15.5,16.5,17.5,18.5,19.5,20.5,21.5,22.5,23.5,24.5,25.5,26.5,27.5,28.5,29.5,30.5,31.5,32.5,33.5,34.5)
axis(1,at=labs,unique(combo_patch_dir_ordered),las=2);
abline(v=match_labs[2:length(match_labs)]);

K <- 4
barplot(t(new.K.4$omega[order(elevation_metadata),]),col=cbbPalette[2:(K+1)],axisnames=F,space=0,border=NA,
        main="",las=1,ylim=c(0,1),cex.axis=1.5,cex.main=1.4)
title(main=paste("Taxonomic Assemblage Motifs, K =",K))
combo_patch_dir = paste0(forest_patch_label,"_",east_west_dir);
combo_patch_dir_ordered = combo_patch_dir[order(east_west_dir)];
match_labs=match(unique(combo_patch_dir_ordered),combo_patch_dir_ordered);
match_labs_suffix=c(match_labs[2:length(unique(combo_patch_dir_ordered))],35);
match_labs_prefix=match_labs[1:(length(unique(combo_patch_dir_ordered)))];
labs=match_labs_prefix + 0.5*(match_labs_suffix - match_labs_prefix);
labs=c(0.5,1.5,2.5,3.5,4.5,5.5,6.5,7.5,8.5,9.5,10.5,11.5,12.5,13.5,14.5,15.5,16.5,17.5,18.5,19.5,20.5,21.5,22.5,23.5,24.5,25.5,26.5,27.5,28.5,29.5,30.5,31.5,32.5,33.5,34.5)
axis(1,at=labs,unique(combo_patch_dir_ordered),las=2);
abline(v=match_labs[2:length(match_labs)]);
```

The Structure plot for K=3 is the same and for K=4 it is clearly different.  There is no obvious way that new_counts or No zero counts should effect the results of the GoM model except that for No zero counts columns were dropped.  Either the wrong columns are being dropped not as intended or the model is sensative to comlumns being included.  However, becuase the K=3 result is so similar, the wrong columns problem is not likely the same.  I am not sure of the answer, but we should investigate before moving forward.  

We look at the theta matrix for the previous fit for the features with all zero abundances across all the forest spots. Ideally the theta values should be pretty small for all these features as they do not contribute much to the clustering.

```{r echo=TRUE, eval=TRUE}
K.3$theta[which(colSums(new_counts) == 0),]
K.4$theta[which(colSums(new_counts) == 0),]
```

Note that for $K=4$, the fourth cluster indeed has much higher theta values than rest for $K=4$ and for $K=3$, they are pretty similar. I think that is the reason why we see the clustering pattern to remain the same from K.3 to new.K.3 but not from K.4 to new.K.4. I think it is because of some discrepancy in the prior setting most likely. You can see all these features have same theta as should be the case as they all contribute equally (all 0's), but the difference in theta across the clusters is because of prior settings and that seems to be driving the differences majorly.

```{r}
sessionInfo()
```


