---
title: "Network visualization"
author: "Kushal K Dey"
date: "March 30, 2016"
output: html_document
---

## Overview

We present an overview of the network visualization tools available in R.

```{r echo=TRUE, eval=FALSE}
install.packages("igraph",repos = "http://cran.us.r-project.org")
install.packages("network", repos = "http://cran.us.r-project.org")
install.packages("sna", repos = "http://cran.us.r-project.org")
install.packages("ndtv", repos = "http://cran.us.r-project.org")
install.packages("RColorBrewer",repos = "http://cran.us.r-project.org")
install.packages("extrafont",repos = "http://cran.us.r-project.org")
```

We load the libraries

```{r echo=TRUE, eval=TRUE, warning=FALSE, message=FALSE}
library(igraph)
library(network)
library(sna)
library(ndtv)
library(RColorBrewer)
library(extrafont)
```

Load an example network data

```{r echo=TRUE, eval=TRUE, message=FALSE, warning=FALSE}
nodes <- read.csv("../external_data/Dataset1-Media-Example-NODES.csv", header=T, as.is=T)
links <- read.csv("../external_data/Dataset1-Media-Example-EDGES.csv", header=T, as.is=T)
```

A brief look at the data

```{r}
head(nodes)
head(links)
nrow(nodes); length(unique(nodes$id))
nrow(links); nrow(unique(links[,c("from", "to")]))

links <- aggregate(links[,3], links[,-3], sum)
links <- links[order(links$from, links$to),]
colnames(links)[4] <- "weight"
rownames(links) <- NULL
```

Load an example matrix data

```{r}
nodes2 <- read.csv("../external_data/Dataset2-Media-User-Example-NODES.csv", header=T, as.is=T)
links2 <- read.csv("../external_data/Dataset2-Media-User-Example-EDGES.csv", header=T, row.names=1)

links2 <- as.matrix(links2)
dim(links2)
dim(nodes2)

```

We first build a graph data frame using the CRAN package \CRANpkg{igraph}. 

In the input arguments of this package,

\begin{itemize}
\item d describes the edges of the network. Its first two columns are the IDs of the source and the target node for each edge. The following columns are edge attributes (weight, type, label, or
anything else).
\item vertices starts with a column of node IDs. Any following columns are interpreted as node
attributes.
\end{itemize}

```{r}
library(igraph)
net <- graph.data.frame(links, nodes, directed=T)
net
```

The description of an igraph object starts with four letters: \newline

D or U, for a directed or undirected graph
N for a named graph (where nodes have a name attribute)
W for a weighted graph (where edges have a weight attribute)
B for a bipartite (two-mode) graph (where nodes have a type attribute)

(g/c) - graph-level character attribute
(v/c) - vertex-level character attribute
(e/n) - edge-level numeric attribute

```{r}
E(net)       # The edges of the "net" object
V(net)       # The vertices of the "net" object
E(net)$type  # Edge attribute "type"
V(net)$media # Vertex attribute "media"

# You can also manipulate the network matrix directly:
net[1,]
net[5,7]

plot(net)
```

This plot is not at all pretty.

Lets remove the multiple edges and the loops.

```{r echo=TRUE, eval=TRUE}
net <- simplify(net, remove.multiple = F, remove.loops = T) 
plot(net, edge.arrow.size=.4,vertex.label=NA)
```

Load the colors we shall use for the network building.

```{r}
display.brewer.all()
display.brewer.pal(8, "Set3")
display.brewer.pal(6, "Blues")
```

Lets look at the different types of fonts.

```{r}
# Import system fonts - may take a while, so DO NOT run this during the workshop.
#font_import() 
fonts() # See what font families are available to you now.

plot(net, vertex.size=30, edge.arrow.size=.4)
plot(net, vertex.size=30, edge.arrow.size=.4, 
     vertex.label.family="Arial Black",
     vertex.color="cyan",
     vertex.size=16,
     vertex.label.font=2,
     vertex.label.cex	=0.5,
     edge.color="black",
     frame=TRUE,
     edge.curved=0.1,
     vertex.label=V(net)$media)
```

We add more attributes to the network visualization.

```{r}
# Generate colors base on media type:
colrs <- c("gray50", "tomato", "gold")
V(net)$color <- colrs[V(net)$media.type]

# Compute node degrees (#links) and use that to set node size:
deg <- igraph::degree(net, mode="all")
V(net)$size <- deg*3
# We could also use the audience size value:
V(net)$size <- V(net)$audience.size*0.6

# The labels are currently node IDs.
# Setting them to NA will render no labels:
V(net)$label <- NA

# Set edge width based on weight:
E(net)$width <- E(net)$weight/6

#change arrow size and edge color:
E(net)$arrow.size <- .5
E(net)$edge.color <- "gray80"
E(net)$width <- 1+E(net)$weight/12
plot(net) 

```

We label the nodes by their type labels.

```{r}
plot(net) 
legend(x=-1.5, y=-1.1, c("Newspaper","Television", "Online News"), pch=21, col="#777777", pt.bg=colrs, pt.cex=2, cex=.8, 
bty="n", ncol=1)
```

In case, we do not want the node balls to show.

```{r}
plot(net, vertex.shape="none", vertex.label=V(net)$media, 
     vertex.label.font=2, vertex.label.color="gray40",
     vertex.label.cex=.7, edge.color="gray85")
```

```{r}
col <- rep("grey40", vcount(net))
col[V(net)$media=="Wall Street Journal"] <- "#ff5100"
neigh.nodes <- neighbors(net, V(net)[media=="Wall Street Journal"], mode="out")
col[neigh.nodes] <- "#ff9d00"
plot(net, vertex.color=col)
plot(net, mark.groups=c(1,4,5,8), mark.col="#C5E5E7", mark.border=NA)
plot(net, mark.groups=list(c(1,4,5,8), c(15:17)), 
          mark.col=c("#C5E5E7","#ECD89A"), mark.border=NA)
```

We can use the heatmap representation of the network.

```{r}
netm <- get.adjacency(net, attr="weight", sparse=F)
colnames(netm) <- V(net)$media
rownames(netm) <- V(net)$media

palf <- colorRampPalette(c("gold", "dark orange")) 
heatmap(netm[,17:1], Rowv = NA, Colv = NA, col = palf(100), 
        scale="none", margins=c(10,10) )

dd <- degree.distribution(net, cumulative=T)
plot(dd, pch=19, cex=1, col="orange", xlab="Degree", ylab="Cumulative Frequency")

net2 <- graph.incidence(links2)
table(E(net2)$type)

plot(net2, vertex.label=NA)
```

We now use the network package to determine the network.

```{r}
library(network)
net3 <- network(links,  vertex.attr=nodes, matrix.type="edgelist", 
                loops=F, multiple=F, ignore.eval = F)

net3[,]
net3 %n% "net.name" <- "Media Network" #  network attribute
net3 %v% "media"    # Node attribute
net3 %e% "type"     # Node attribute

net3 %v% "col" <- c("gray70", "tomato", "gold")[net3 %v% "media.type"]
```

