---
title: "Null models for GoM model fit"
author: "Kushal K Dey"
date: "March 14, 2016"
output: 
  html_document:
    css: floating-toc.css
    toc: true
---

```{r echo=TRUE, eval=TRUE, message=FALSE, warning=FALSE}
library(CountClust)
library(picante)
```

## Overview

The null models are an useful and widely used technique to test for the randomization of ecological presenece absence data. We apply a similar mechanism now for the GoM model fits. 

```{r echo=TRUE, eval=TRUE}
data=read.csv("../external_data/Himalayan_grid_matrix.csv",header=TRUE);
counts=as.matrix(data[,-1]);
rownames(counts)=data[,1];
new_data1 <- data.frame(read.csv('../external_data/MohanGrids.csv'));
new_data2 <- data.frame(read.csv('../external_data/MohanGrids2.csv'));
bird_species <- union(as.vector(colnames(counts)), union(as.vector(new_data1[,1]), as.vector(new_data2[,1])));
new_data <- matrix(0,dim(counts)[1]+3,length(bird_species));
new_data[1,match(new_data1[,1],bird_species)]=new_data1[,2];
new_data[2,match(new_data1[,1],bird_species)]=new_data1[,3];
new_data[3,match(new_data2[,1],bird_species)]=new_data2[,2];
new_data[4:(dim(counts)[1]+3),match(colnames(counts),bird_species)]=counts;
new_counts <- as.matrix(new_data);
rownames(new_counts) <- c(c("U1","U2","MA1"),rownames(counts));
colnames(new_counts) <- bird_species;
new_counts <- new_counts[-(1:3),];

```

We load the metadata

```{r echo=TRUE, eval=TRUE}
metadata=read.csv("../external_data/Himalayan_grid_metadata.csv",header=TRUE);
elevation_metadata=metadata$Elevation;

east_west_dir = metadata$WorE;
```

We use the `randomizeMatrix()` function due to the package `picante` for applying null models and testing for the efficcacy of the clustering we use. 

## GoM model fit (maptpx) - original data 

```{r echo=TRUE, eval=TRUE}
topics_clus <- FitGoM(new_counts, K=2, tol=0.1);

K <- 2
barplot(t(topics_clus[[1]]$omega[order(elevation_metadata),]),col=2:(K+1),axisnames=F,space=0,border=NA,main="",las=1,ylim=c(0,1),cex.axis=1.5,cex.main=1.4, xlab="elevation")
title(main=paste("Structure Plot topic proportions,k=",K))
combo_patch_dir = paste0(round(elevation_metadata));
combo_patch_dir_ordered = combo_patch_dir[order(elevation_metadata)];

match_labs=match(unique(combo_patch_dir_ordered),combo_patch_dir_ordered);
match_labs_suffix=c(match_labs[2:length(unique(combo_patch_dir_ordered))],35);
match_labs_prefix=match_labs[1:(length(unique(combo_patch_dir_ordered)))];
labs=match_labs_prefix + 0.5*(match_labs_suffix - match_labs_prefix);
  
axis(1,at=labs,unique(combo_patch_dir_ordered),las=2);
abline(v=match_labs[2:length(match_labs)]);

```

## Randomized data + maptpx fit (K=2)

```{r echo=TRUE, eval=TRUE}
null_counts <- picante::randomizeMatrix(new_counts, null.model = "richness", iterations=1000);

topics_null <- FitGoM(null_counts, K=2, tol=0.1);

K <- 2
barplot(t(topics_null[[1]]$omega[order(elevation_metadata),]),col=2:(K+1),axisnames=F,space=0,border=NA,main="",las=1,ylim=c(0,1),cex.axis=1.5,cex.main=1.4, xlab="elevation")
title(main=paste("Structure Plot topic proportions,k=",K))
combo_patch_dir = paste0(round(elevation_metadata));
combo_patch_dir_ordered = combo_patch_dir[order(elevation_metadata)];

match_labs=match(unique(combo_patch_dir_ordered),combo_patch_dir_ordered);
match_labs_suffix=c(match_labs[2:length(unique(combo_patch_dir_ordered))],35);
match_labs_prefix=match_labs[1:(length(unique(combo_patch_dir_ordered)))];
labs=match_labs_prefix + 0.5*(match_labs_suffix - match_labs_prefix);
  
axis(1,at=labs,unique(combo_patch_dir_ordered),las=2);
abline(v=match_labs[2:length(match_labs)]);


```

## Null model evaluation

We use the `nullmodel_GoM()` functionality in the CountClust package to obtain statistical measures of 
how much the observed counts matrix differs from null randomized matrices.

Under the default, we generate 100 randomized matrices and compare the observed
matrix log Bayes factor with that of the randomized designs.

### Trialswap method

We first use the `trialswap` method. 

```{r echo=TRUE, eval=TRUE, message=FALSE, warning=FALSE}
out <- nullmodel_GoM(new_counts, K=2, null.model = "trialswap")
print(out$pval)
```

### Richness method

Then we use the `richness` method.

```{r echo=TRUE, eval=TRUE, message=FALSE, warning=FALSE}
out <- nullmodel_GoM(new_counts, K=2, null.model = "richness")
print(out$pval)
```

### Frequency method

Consider now the `frequency` method.

```{r echo=TRUE, eval=TRUE, message=FALSE, warning=FALSE}
out <- nullmodel_GoM(new_counts, K=2, null.model = "frequency")
print(out$pval)
```

### Independent Swap method

Finally we consider the `independentswap` method.

```{r echo=TRUE, eval=TRUE, message=FALSE, warning=FALSE}
out <- nullmodel_GoM(new_counts, K=2, null.model = "independentswap")
print(out$pval)
```


## Conclusion

It seems apart from trialswap method, the p-value of the log Bayes factor for the observed data (K=2 vs null) against the distribution of the log Bayes factors for all the randomized data (K=2 vs null) is close to 0. Even for trialswap method it is pretty small, which validates the clustering and solidifies the belief in the clustering.

