---
title: 'Ordered topic model: Alex Birds data'
author: "Kushal K Dey + Alex White"
date: "April 18, 2016"
output: html_document
---

## Overview

We use the `ordtpx` package to perform ordered topic model on the bird abundance data from Trevor Price's lab where the bird species are ordered by increasing bodymass/ beak size. For this version of topic models, we assume a multi-resolution prior on the bird species, ordered by bodymasses. For the first application of ordered topic models, we have left out nugget effects from the analysis. 

Here we show ordtpx results under high smoothness and low smoothness conditions and also standard topic model (maptpx) results.

```{r echo=TRUE, eval=TRUE, message=FALSE, warning=FALSE}
library(devtools)
#install_github('kkdey/ordtpx')
library(ordtpx)
library(maptpx)
library(plotrix)
```


```{r echo=TRUE, eval=TRUE}

data=read.csv("Himalayan_grid_matrix.csv",header=TRUE);

counts=as.matrix(data[,-1]);

rownames(counts)=data[,1];

new_data1 <- data.frame(read.csv('MohanGrids.csv'));
new_data2 <- data.frame(read.csv('MohanGrids2.csv'));

bird_species <- union(as.vector(colnames(counts)), union(as.vector(new_data1[,1]), as.vector(new_data2[,1])));

new_data <- matrix(0,dim(counts)[1]+3,length(bird_species));
new_data[1,match(new_data1[,1],bird_species)]=new_data1[,2];
new_data[2,match(new_data1[,1],bird_species)]=new_data1[,3];
new_data[3,match(new_data2[,1],bird_species)]=new_data2[,2];
new_data[4:(dim(counts)[1]+3),match(colnames(counts),bird_species)]=counts;

new_counts <- as.matrix(new_data);

rownames(new_counts) <- c(c("U1","U2","MA"),rownames(counts));
colnames(new_counts) <- bird_species;

new_counts <- new_counts[,colnames(new_counts[,which(colSums(new_counts) > 0)])]

metadata <- read.csv("gridsmeta.csv",header=T)
metadata <-metadata[c(9,8,1,10,6,2,3,36,7,11,5,33,4,30,14,32,31,12,13,37,15,34,19,38,18,35,17,23,16,24,28,21,26,20,27,25,22,29),]
row.names(metadata) <- metadata$GRID
east_west_dir = metadata$WorE
forest_patch_label <- row.names(metadata)

cbbPalette <- c("#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")

bodymass_data <- read.csv('massKushal.csv');

bodymass_val <- bodymass_data[,2];
bodymass_birds <- bodymass_data[,1];

bodymass_matched <- bodymass_data[match(colnames(new_counts),bodymass_birds),2];
summary(bodymass_matched)

morph <- read.csv(file = "AllDatMerged3.csv", header = T, row.names =1)

morph <- morph[as.vector(colnames(new_counts)),]
morph <- na.omit(morph[,c(1,2,3,4,5,12,47:52)])
morph.log <- log(morph[,1:6])
morph[,7:12][morph[,7:12] == 5] <- 1
morph.gow <- morph[,7:12]
morph <- cbind(morph.log, morph[,7:12])

morph.pc <- prcomp(morph, center=T, scale=T)
morph.pc$rotation
summary(morph.pc)

morph.scores <- as.data.frame(morph.pc$x)
morph.scores.1.2 <- morph.scores[,1:2]
morph.scores.1 <- as.data.frame(morph.scores.1.2[,1])
morph.scores.2 <- as.data.frame(morph.scores.1.2[,2])
morph.scores.3 <- as.data.frame(morph.scores[,3])
morph.scores.4 <- as.data.frame(morph.scores[,4])

rownames(morph.scores.1) <- as.vector(rownames(morph))
colnames(morph.scores.1) <- c("PC1")
rownames(morph.scores.2) <- as.vector(rownames(morph))
colnames(morph.scores.2) <- c("PC2")
rownames(morph.scores.3) <- as.vector(rownames(morph))
colnames(morph.scores.3) <- c("PC3")
rownames(morph.scores.4) <- as.vector(rownames(morph))
colnames(morph.scores.4) <- c("PC4")

PC1_val <- as.vector(morph.scores.1[,1])
PC_birds <- as.vector(rownames(morph.scores.1))

PC1_matched <- morph.scores.1[match(colnames(new_counts),PC_birds),1];
summary(PC1_matched)

PC2_val <- as.vector(morph.scores.2[,1])
PC_birds <- as.vector(rownames(morph.scores.2))

PC2_matched <- morph.scores.2[match(colnames(new_counts),PC_birds),1];
summary(PC2_matched)

PC3_val <- as.vector(morph.scores.3[,1])
PC_birds <- as.vector(rownames(morph.scores.3))

PC3_matched <- morph.scores.3[match(colnames(new_counts),PC_birds),1];
summary(PC3_matched)

PC4_val <- as.vector(morph.scores.4[,1])
PC_birds <- as.vector(rownames(morph.scores.4))

PC4_matched <- morph.scores.4[match(colnames(new_counts),PC_birds),1];
summary(PC4_matched)

ordered_counts <- new_counts[,order(PC1_matched, decreasing = FALSE)]

#ordered_counts <- new_counts[,order(PC2_matched, decreasing = FALSE)]

#ordered_counts <- new_counts[,order(PC3_matched, decreasing = FALSE)]

#ordered_counts <- new_counts[,order(PC4_matched, decreasing = FALSE)]

#ordered_counts <- new_counts[,order(bodymass_matched, decreasing = FALSE)];


```

## ordtpx (K=2) [high smoothness]

```{r echo=TRUE, eval=TRUE}
library(ordtpx)
levels <- ceiling(log(dim(ordered_counts)[2])/log(2))+1;

Topic_clus_list <- vector(mode = "list", length = 25)
loglik_clus_list <- vector(mode = "list", length = 25)

scale <- 2
del_beta <- scale*(10^{1:(levels-1)})
Topic_clus <- ord_topics(ordered_counts, K=2, del_beta = del_beta, reflect=FALSE, a_mu=20, b_mu=30, ztree_options=2, tol=0.00000000000000001)
Topic_clus$loglik
theta <- Topic_clus$theta
omega <-Topic_clus$omega
probs <- omega %*% t(theta)
loglik <- Reduce(sum, sapply(1:NROW(ordered_counts), function(i) {
            dmultinom(x = ordered_counts[i,], prob = probs[i,1:304], log = TRUE)
        }) )

loglik <- loglik - (10^4)*sum((theta[2:dim(theta)[1],] - theta[1:(dim(theta)[1]-1)])^2)
loglik
Topic_clus_list[[scale]] <- Topic_clus
loglik_clus_list[[scale]] <- loglik;
#print(scale)
```

### Structure Plot

```{r echo=TRUE, eval=TRUE}
K <- 2
par(mfrow=c(1,1))
{barplot(t(Topic_clus$omega[forest_patch_label,]),col=cbbPalette[1:K],axisnames=F,space=0,border=NA,
        main="",las=1,ylim=c(0,1),cex.axis=1.5,cex.main=1.4)
title(main=paste("Bodymass motif proportions, K =",K))
combo_patch_dir = paste0(forest_patch_label,"_",east_west_dir);
combo_patch_dir_ordered = combo_patch_dir[order(east_west_dir)];
match_labs=match(unique(combo_patch_dir_ordered),combo_patch_dir_ordered);
match_labs_suffix=c(match_labs[2:length(unique(combo_patch_dir_ordered))],38);
match_labs_prefix=match_labs[1:(length(unique(combo_patch_dir_ordered)))];
labs=match_labs_prefix + 0.5*(match_labs_suffix - match_labs_prefix);
labs=c(0.5,1.5,2.5,3.5,4.5,5.5,6.5,7.5,8.5,9.5,10.5,11.5,12.5,13.5,14.5,15.5,16.5,17.5,18.5,19.5,20.5,21.5,22.5,23.5,24.5,25.5,26.5,27.5,28.5,29.5,30.5,31.5,32.5,33.5,34.5,35.5,36.5,37.5)
axis(1,at=labs,unique(combo_patch_dir_ordered),las=2);
ablineclip(v=match_labs[0:length(match_labs)],y1=0,y2=1);}

```

### Topic plots plots

```{r echo=TRUE, eval=TRUE}
par(mfrow=c(1,2))
plot(Topic_clus$theta[,1], type="l")
plot(Topic_clus$theta[,2], type="l")
```

## ordtpx K=3 [high smoothness]

```{r echo=TRUE, eval=TRUE}
del_beta <- scale*(10^{1:(levels-1)})
Topic_clus <- ord_topics(ordered_counts, K=3, del_beta = del_beta, reflect=FALSE, a_mu=20, b_mu=30, ztree_options=2, tol=0.00000000000000001)
docweights <- Topic_clus$omega;
Topic_clus$loglik
theta <- Topic_clus$theta
omega <-Topic_clus$omega
probs <- omega %*% t(theta)
loglik <- Reduce(sum, sapply(1:NROW(ordered_counts), function(i) {
            dmultinom(x = ordered_counts[i,], prob = probs[i,1:304], log = TRUE)
        }) )
loglik <- loglik - (10^4)*sum((theta[2:dim(theta)[1],] - theta[1:(dim(theta)[1]-1)])^2)
loglik
Topic_clus_list <- list(Topic_clus)
```

### Structure Plot

```{r echo=TRUE, eval=TRUE}
K<-3
par(mfrow=c(1,1))
{barplot(t(docweights[forest_patch_label,]),col=cbbPalette[1:K],axisnames=F,space=0,border=NA,
        main="",las=1,ylim=c(0,1),cex.axis=1.5,cex.main=1.4)
title(main=paste("Bodymass motif proportions, K =",K))
combo_patch_dir = paste0(forest_patch_label,"_",east_west_dir);
combo_patch_dir_ordered = combo_patch_dir[order(east_west_dir)];
match_labs=match(unique(combo_patch_dir_ordered),combo_patch_dir_ordered);
match_labs_suffix=c(match_labs[2:length(unique(combo_patch_dir_ordered))],38);
match_labs_prefix=match_labs[1:(length(unique(combo_patch_dir_ordered)))];
labs=match_labs_prefix + 0.5*(match_labs_suffix - match_labs_prefix);
labs=c(0.5,1.5,2.5,3.5,4.5,5.5,6.5,7.5,8.5,9.5,10.5,11.5,12.5,13.5,14.5,15.5,16.5,17.5,18.5,19.5,20.5,21.5,22.5,23.5,24.5,25.5,26.5,27.5,28.5,29.5,30.5,31.5,32.5,33.5,34.5,35.5,36.5,37.5)
axis(1,at=labs,unique(combo_patch_dir_ordered),las=2);
ablineclip(v=match_labs[0:length(match_labs)],y1=0,y2=1);}
```

### Cluster plots

```{r echo=TRUE, eval=TRUE}
par(mfrow=c(1,3))
plot(Topic_clus$theta[1:304,1], type="l")
plot(Topic_clus$theta[1:304,2], type="l")
plot(Topic_clus$theta[1:304,3], type="l")

```

## ordtpx (K=2) [low smoothness]

```{r echo=TRUE, eval=TRUE}
library(ordtpx)
levels <- ceiling(log(dim(ordered_counts)[2])/log(2))+1;

Topic_clus_list <- vector(mode = "list", length = 25)
loglik_clus_list <- vector(mode = "list", length = 25)

scale <- 2
del_beta <- scale*(1.2^{1:(levels-1)})
Topic_clus <- ord_topics(ordered_counts, K=2, del_beta = del_beta, reflect=FALSE, a_mu=20, b_mu=30, ztree_options=2, tol=0.00000000000000001)
Topic_clus$loglik
theta <- Topic_clus$theta
omega <-Topic_clus$omega
probs <- omega %*% t(theta)
loglik <- Reduce(sum, sapply(1:NROW(ordered_counts), function(i) {
            dmultinom(x = ordered_counts[i,], prob = probs[i,1:304], log = TRUE)
        }) )

loglik <- loglik - (10^4)*sum((theta[2:dim(theta)[1],] - theta[1:(dim(theta)[1]-1)])^2)
loglik
Topic_clus_list[[scale]] <- Topic_clus
loglik_clus_list[[scale]] <- loglik;
#print(scale)
```

### Structure Plot

```{r echo=TRUE, eval=TRUE}
K<-2
par(mfrow=c(1,1))
{barplot(t(Topic_clus$omega[forest_patch_label,]),col=cbbPalette[1:K],axisnames=F,space=0,border=NA,
        main="",las=1,ylim=c(0,1),cex.axis=1.5,cex.main=1.4)
title(main=paste("Bodymass motif proportions, K =",K))
combo_patch_dir = paste0(forest_patch_label,"_",east_west_dir);
combo_patch_dir_ordered = combo_patch_dir[order(east_west_dir)];
match_labs=match(unique(combo_patch_dir_ordered),combo_patch_dir_ordered);
match_labs_suffix=c(match_labs[2:length(unique(combo_patch_dir_ordered))],38);
match_labs_prefix=match_labs[1:(length(unique(combo_patch_dir_ordered)))];
labs=match_labs_prefix + 0.5*(match_labs_suffix - match_labs_prefix);
labs=c(0.5,1.5,2.5,3.5,4.5,5.5,6.5,7.5,8.5,9.5,10.5,11.5,12.5,13.5,14.5,15.5,16.5,17.5,18.5,19.5,20.5,21.5,22.5,23.5,24.5,25.5,26.5,27.5,28.5,29.5,30.5,31.5,32.5,33.5,34.5,35.5,36.5,37.5)
axis(1,at=labs,unique(combo_patch_dir_ordered),las=2);
ablineclip(v=match_labs[0:length(match_labs)],y1=0,y2=1);}
```

### Topic plots plots

```{r echo=TRUE, eval=TRUE}
par(mfrow=c(1,2))
plot(Topic_clus$theta[,1], type="l")
plot(Topic_clus$theta[,2], type="l")
```

## ordtpx K=3 [low smoothness]

```{r echo=TRUE, eval=TRUE}
del_beta <- scale*(1.02^{1:(levels-1)})
Topic_clus <- ord_topics(ordered_counts, K=3, del_beta = del_beta, reflect=FALSE, a_mu=20, b_mu=30, ztree_options=2, tol=0.00000000000000001)
docweights <- Topic_clus$omega;
Topic_clus$loglik
theta <- Topic_clus$theta
omega <-Topic_clus$omega
probs <- omega %*% t(theta)
loglik <- Reduce(sum, sapply(1:NROW(ordered_counts), function(i) {
            dmultinom(x = ordered_counts[i,], prob = probs[i,1:304], log = TRUE)
        }) )
loglik <- loglik - (10^4)*sum((theta[2:dim(theta)[1],] - theta[1:(dim(theta)[1]-1)])^2)
loglik
Topic_clus_list <- list(Topic_clus)
```

### Structure Plot

```{r echo=TRUE, eval=TRUE}
par(mfrow=c(1,1))
K<-3
{barplot(t(docweights[forest_patch_label,]),col=cbbPalette[1:K],axisnames=F,space=0,border=NA,
        main="",las=1,ylim=c(0,1),cex.axis=1.5,cex.main=1.4)
title(main=paste("Bodymass motif proportions, K =",K))
combo_patch_dir = paste0(forest_patch_label,"_",east_west_dir);
combo_patch_dir_ordered = combo_patch_dir[order(east_west_dir)];
match_labs=match(unique(combo_patch_dir_ordered),combo_patch_dir_ordered);
match_labs_suffix=c(match_labs[2:length(unique(combo_patch_dir_ordered))],38);
match_labs_prefix=match_labs[1:(length(unique(combo_patch_dir_ordered)))];
labs=match_labs_prefix + 0.5*(match_labs_suffix - match_labs_prefix);
labs=c(0.5,1.5,2.5,3.5,4.5,5.5,6.5,7.5,8.5,9.5,10.5,11.5,12.5,13.5,14.5,15.5,16.5,17.5,18.5,19.5,20.5,21.5,22.5,23.5,24.5,25.5,26.5,27.5,28.5,29.5,30.5,31.5,32.5,33.5,34.5,35.5,36.5,37.5)
axis(1,at=labs,unique(combo_patch_dir_ordered),las=2);
ablineclip(v=match_labs[0:length(match_labs)],y1=0,y2=1);}
```

### Cluster plots

```{r echo=TRUE, eval=TRUE}
par(mfrow=c(1,3))
plot(Topic_clus$theta[1:304,1], type="l")
plot(Topic_clus$theta[1:304,2], type="l")
plot(Topic_clus$theta[1:304,3], type="l")

```


## Fitting standard topic model (K=2)

```{r echo=TRUE, eval=TRUE}
Topic_clus <- maptpx::topics(ordered_counts, K=2, tol=0.0001)
```

### Structure Plot

```{r echo=TRUE, eval=TRUE}
K <- 2
par(mfrow=c(1,1))
{barplot(t(Topic_clus$omega[forest_patch_label,]),col=cbbPalette[1:K],axisnames=F,space=0,border=NA,
        main="",las=1,ylim=c(0,1),cex.axis=1.5,cex.main=1.4)
title(main=paste("Taxonomic motif proportions, K =",K))
combo_patch_dir = paste0(forest_patch_label,"_",east_west_dir);
combo_patch_dir_ordered = combo_patch_dir[order(east_west_dir)];
match_labs=match(unique(combo_patch_dir_ordered),combo_patch_dir_ordered);
match_labs_suffix=c(match_labs[2:length(unique(combo_patch_dir_ordered))],38);
match_labs_prefix=match_labs[1:(length(unique(combo_patch_dir_ordered)))];
labs=match_labs_prefix + 0.5*(match_labs_suffix - match_labs_prefix);
labs=c(0.5,1.5,2.5,3.5,4.5,5.5,6.5,7.5,8.5,9.5,10.5,11.5,12.5,13.5,14.5,15.5,16.5,17.5,18.5,19.5,20.5,21.5,22.5,23.5,24.5,25.5,26.5,27.5,28.5,29.5,30.5,31.5,32.5,33.5,34.5,35.5,36.5,37.5)
axis(1,at=labs,unique(combo_patch_dir_ordered),las=2);
ablineclip(v=match_labs[0:length(match_labs)],y1=0,y2=1);}

```

### Cluster plots

```{r echo=TRUE, eval=TRUE}
par(mfrow=c(1,2))
plot(Topic_clus$theta[,1], type="l")
plot(Topic_clus$theta[,2], type="l")
```

## Fitting standard topic model (K=3)

```{r echo=TRUE, eval=TRUE}
Topic_clus <- maptpx::topics(ordered_counts, K=3, tol=0.0001)
docweights <- Topic_clus$omega;
```

### Structure Plot

```{r echo=TRUE, eval=TRUE}
par(mfrow=c(1,1))
K<-3
{barplot(t(docweights[forest_patch_label,]),col=cbbPalette[1:K],axisnames=F,space=0,border=NA,
        main="",las=1,ylim=c(0,1),cex.axis=1.5,cex.main=1.4)
title(main=paste("Taxonomic motif proportions, K =",K))
combo_patch_dir = paste0(forest_patch_label,"_",east_west_dir);
combo_patch_dir_ordered = combo_patch_dir[order(east_west_dir)];
match_labs=match(unique(combo_patch_dir_ordered),combo_patch_dir_ordered);
match_labs_suffix=c(match_labs[2:length(unique(combo_patch_dir_ordered))],38);
match_labs_prefix=match_labs[1:(length(unique(combo_patch_dir_ordered)))];
labs=match_labs_prefix + 0.5*(match_labs_suffix - match_labs_prefix);
labs=c(0.5,1.5,2.5,3.5,4.5,5.5,6.5,7.5,8.5,9.5,10.5,11.5,12.5,13.5,14.5,15.5,16.5,17.5,18.5,19.5,20.5,21.5,22.5,23.5,24.5,25.5,26.5,27.5,28.5,29.5,30.5,31.5,32.5,33.5,34.5,35.5,36.5,37.5)
axis(1,at=labs,unique(combo_patch_dir_ordered),las=2);
ablineclip(v=match_labs[0:length(match_labs)],y1=0,y2=1);}
```

### Cluster plots

```{r echo=TRUE, eval=TRUE}
par(mfrow=c(1,3))
plot(Topic_clus$theta[,1], type="l")
plot(Topic_clus$theta[,2], type="l")
plot(Topic_clus$theta[,3], type="l")

```


## Conclusions 

Under low smoothness, the ordtpx results closely match with standard topic model results. But the high smoothness case creates topics which are smoothly varying across the bird species bodymass and the clusters formed are different (specially for K=3 case) compared to the ordtpx with low smoothness and standard topic model fit.

