---
title: 'Alex Bird Data: sliding window binning by log bodymass'
author: "Kushal K Dey"
date: "March 3, 2016"
output: 
  html_document:
    css: floating-toc.css
    toc: true
---

## Background

We perform a topic model on the counts obtained from running window or sliding window based on log bodymass/ bodymass information. 

## Loading bird abundance Data

```{r echo=TRUE, eval=TRUE}
library(maptpx)
data=read.csv("../external_data/Himalayan_grid_matrix.csv",header=TRUE);

counts=as.matrix(data[,-1]);

rownames(counts)=data[,1];

new_data1 <- data.frame(read.csv('../external_data/MohanGrids.csv'));
new_data2 <- data.frame(read.csv('../external_data/MohanGrids2.csv'));

bird_species <- union(as.vector(colnames(counts)), union(as.vector(new_data1[,1]), as.vector(new_data2[,1])));

new_data <- matrix(0,dim(counts)[1]+3,length(bird_species));
new_data[1,match(new_data1[,1],bird_species)]=new_data1[,2];
new_data[2,match(new_data1[,1],bird_species)]=new_data1[,3];
new_data[3,match(new_data2[,1],bird_species)]=new_data2[,2];
new_data[4:(dim(counts)[1]+3),match(colnames(counts),bird_species)]=counts;

new_counts <- as.matrix(new_data);

rownames(new_counts) <- c(c("U1","U2","MA1"),rownames(counts));
colnames(new_counts) <- bird_species;
new_counts <- new_counts[-(1:3),];
```

### Loading metadata

Next we load the required metadata for forest spots.

```{r echo=TRUE, eval=TRUE}
metadata=read.csv("../external_data/Himalayan_grid_metadata.csv",header=TRUE);
elevation_metadata=metadata$Elevation[match(rownames(new_counts),metadata[,1])];
east_west_dir = metadata$WorE[match(rownames(new_counts),metadata[,1])];

```

### Loading bodymass data

Now we load the bodymass data for the bird species. 

```{r echo=TRUE, eval=TRUE}
bodymass_data <- read.csv('../external_data/massKushal.csv');

bodymass_val <- bodymass_data[,2];
bodymass_birds <- bodymass_data[,1];

bodymass_matched <- bodymass_data[match(colnames(new_counts),bodymass_birds),2];
summary(bodymass_matched)

```

Order the bird species data based on bodymass 

```{r echo=TRUE, eval=TRUE}
ordered_counts <- new_counts[,order(bodymass_matched, decreasing = FALSE)];
```

Look at the log scale bodymass and then obtain histogram breaks. Group counts for bird species falling in each break.

## Sliding window: n_breaks: 10

```{r echo=TRUE, eval=TRUE}
n_breaks <- 10;
bodymass_ordered <- bodymass_matched[order(bodymass_matched, decreasing = FALSE)];
histo <- hist(log(bodymass_ordered+1), breaks=n_breaks)
  #histo$breaks

binned_data = NULL
histo$breaks <- histo$breaks[-1];


for(l in 1:length(histo$breaks))
{
  if(l==1){
      indices <- which(log(bodymass_ordered+1) < histo$breaks[l])
      if(length(indices)==0){ out <- rep(0,dim(ordered_counts)[1])}
      if(length(indices)==1){ out <- ordered_counts[,indices]}
      if(length(indices)> 1){ out <- rowSums(ordered_counts[,indices])}
  }
  if(l > 1){
      indices <- which(log(bodymass_ordered+1) < histo$breaks[l] & log(bodymass_ordered+1) > histo$breaks[(l-1)])
      if(length(indices)==0){ out <- rep(0,dim(ordered_counts)[1])}
      if(length(indices)==1){ out <- ordered_counts[,indices]}
      if(length(indices)> 1){ out <- rowSums(ordered_counts[,indices])}
  }
    binned_data <- cbind(binned_data, out);
}


```

We create sliding window counts now.

```{r echo=TRUE, eval=TRUE}
sliding_binned_data <- t(apply(binned_data, 1, function(x) zoo::rollsum(x, 5)));
```

### Topic model fit 

#### K=2

```{r echo=TRUE, eval=TRUE}
Topic_clus <- topics(sliding_binned_data, K=2)
K=2
barplot(t(Topic_clus$omega[order(elevation_metadata),]),col=2:(K+1),axisnames=F,space=0,border=NA,main="",las=1,ylim=c(0,1),
          cex.axis=1.5,cex.main=1.4)
  title(main=paste("Structure Plot,k=",K,"bins=",n_breaks))
  combo_patch_dir = paste0(elevation_metadata,"_",east_west_dir);
  combo_patch_dir_ordered = combo_patch_dir[order(elevation_metadata)];

  match_labs=match(unique(combo_patch_dir_ordered),combo_patch_dir_ordered);
  match_labs_suffix=c(match_labs[2:length(unique(combo_patch_dir_ordered))],35);
  match_labs_prefix=match_labs[1:(length(unique(combo_patch_dir_ordered)))];
  labs=match_labs_prefix + 0.5*(match_labs_suffix - match_labs_prefix);
  
  axis(1,at=labs,unique(combo_patch_dir_ordered),las=2);
  abline(v=match_labs[2:length(match_labs)]);
  

```

####  K=3

```{r echo=TRUE, eval=TRUE}
Topic_clus <- topics(sliding_binned_data, K=3)
K=3
east_west_elevation = paste0(metadata$WorE, "_", metadata$Elevation);

index1 <- which(metadata$WorE=="E");
index2 <- which(metadata$WorE=="W");
elevation1 <- metadata$Elevation[index1]; elevation2 <- metadata$Elevation[index2];
index_WE <- c(index1[order(elevation1)], index2[order(elevation2)]);

 K <- 3
    barplot(t(Topic_clus$omega[index_WE,]),col=2:(K+1),axisnames=F,space=0,border=NA,main=paste("No. of clusters=",3),las=1,ylim=c(0,1),cex.axis=1.5,cex.main=1.4)

    combo_patch_dir = paste0(east_west_elevation);
    combo_patch_dir_ordered = combo_patch_dir[index_WE];

  match_labs=match(unique(combo_patch_dir_ordered),combo_patch_dir_ordered);
  match_labs_suffix=c(match_labs[2:length(unique(combo_patch_dir_ordered))],35);
  match_labs_prefix=match_labs[1:(length(unique(combo_patch_dir_ordered)))];
  labs=match_labs_prefix + 0.5*(match_labs_suffix - match_labs_prefix);
  
  axis(1,at=labs,unique(combo_patch_dir_ordered),las=2);
  abline(v=match_labs[2:length(match_labs)]);
```


## Sliding window: n_breaks: 30

```{r echo=TRUE, eval=TRUE}
n_breaks <- 30;
bodymass_ordered <- bodymass_matched[order(bodymass_matched, decreasing = FALSE)];
histo <- hist(log(bodymass_ordered+1), breaks=n_breaks)
  #histo$breaks

binned_data = NULL
histo$breaks <- histo$breaks[-1];


for(l in 1:length(histo$breaks))
{
  if(l==1){
      indices <- which(log(bodymass_ordered+1) < histo$breaks[l])
      if(length(indices)==0){ out <- rep(0,dim(ordered_counts)[1])}
      if(length(indices)==1){ out <- ordered_counts[,indices]}
      if(length(indices)> 1){ out <- rowSums(ordered_counts[,indices])}
  }
  if(l > 1){
      indices <- which(log(bodymass_ordered+1) < histo$breaks[l] & log(bodymass_ordered+1) > histo$breaks[(l-1)])
      if(length(indices)==0){ out <- rep(0,dim(ordered_counts)[1])}
      if(length(indices)==1){ out <- ordered_counts[,indices]}
      if(length(indices)> 1){ out <- rowSums(ordered_counts[,indices])}
  }
    binned_data <- cbind(binned_data, out);
}


```

We create sliding window counts now.

```{r echo=TRUE, eval=TRUE}
sliding_binned_data <- t(apply(binned_data, 1, function(x) zoo::rollsum(x, 5)));
```

### Topic model fit 

#### K=2

```{r echo=TRUE, eval=TRUE}
Topic_clus <- topics(sliding_binned_data, K=2)
K=2
barplot(t(Topic_clus$omega[order(elevation_metadata),]),col=2:(K+1),axisnames=F,space=0,border=NA,main="",las=1,ylim=c(0,1),
          cex.axis=1.5,cex.main=1.4)
  title(main=paste("Structure Plot,k=",K,"bins=",n_breaks))
  combo_patch_dir = paste0(elevation_metadata,"_",east_west_dir);
  combo_patch_dir_ordered = combo_patch_dir[order(elevation_metadata)];

  match_labs=match(unique(combo_patch_dir_ordered),combo_patch_dir_ordered);
  match_labs_suffix=c(match_labs[2:length(unique(combo_patch_dir_ordered))],35);
  match_labs_prefix=match_labs[1:(length(unique(combo_patch_dir_ordered)))];
  labs=match_labs_prefix + 0.5*(match_labs_suffix - match_labs_prefix);
  
  axis(1,at=labs,unique(combo_patch_dir_ordered),las=2);
  abline(v=match_labs[2:length(match_labs)]);
  

```

####  K=3

```{r echo=TRUE, eval=TRUE}
Topic_clus <- topics(sliding_binned_data, K=3)
K=3
east_west_elevation = paste0(metadata$WorE, "_", metadata$Elevation);

index1 <- which(metadata$WorE=="E");
index2 <- which(metadata$WorE=="W");
elevation1 <- metadata$Elevation[index1]; elevation2 <- metadata$Elevation[index2];
index_WE <- c(index1[order(elevation1)], index2[order(elevation2)]);

 K <- 3
    barplot(t(Topic_clus$omega[index_WE,]),col=2:(K+1),axisnames=F,space=0,border=NA,main=paste("No. of clusters=",3),las=1,ylim=c(0,1),cex.axis=1.5,cex.main=1.4)

    combo_patch_dir = paste0(east_west_elevation);
    combo_patch_dir_ordered = combo_patch_dir[index_WE];

  match_labs=match(unique(combo_patch_dir_ordered),combo_patch_dir_ordered);
  match_labs_suffix=c(match_labs[2:length(unique(combo_patch_dir_ordered))],35);
  match_labs_prefix=match_labs[1:(length(unique(combo_patch_dir_ordered)))];
  labs=match_labs_prefix + 0.5*(match_labs_suffix - match_labs_prefix);
  
  axis(1,at=labs,unique(combo_patch_dir_ordered),las=2);
  abline(v=match_labs[2:length(match_labs)]);
```

## Sliding window: n_breaks: 50

```{r echo=TRUE, eval=TRUE}
n_breaks <- 50;
bodymass_ordered <- bodymass_matched[order(bodymass_matched, decreasing = FALSE)];
histo <- hist(log(bodymass_ordered+1), breaks=n_breaks)
  #histo$breaks

binned_data = NULL
histo$breaks <- histo$breaks[-1];


for(l in 1:length(histo$breaks))
{
  if(l==1){
      indices <- which(log(bodymass_ordered+1) < histo$breaks[l])
      if(length(indices)==0){ out <- rep(0,dim(ordered_counts)[1])}
      if(length(indices)==1){ out <- ordered_counts[,indices]}
      if(length(indices)> 1){ out <- rowSums(ordered_counts[,indices])}
  }
  if(l > 1){
      indices <- which(log(bodymass_ordered+1) < histo$breaks[l] & log(bodymass_ordered+1) > histo$breaks[(l-1)])
      if(length(indices)==0){ out <- rep(0,dim(ordered_counts)[1])}
      if(length(indices)==1){ out <- ordered_counts[,indices]}
      if(length(indices)> 1){ out <- rowSums(ordered_counts[,indices])}
  }
    binned_data <- cbind(binned_data, out);
}


```

We create sliding window counts now.

```{r echo=TRUE, eval=TRUE}
sliding_binned_data <- t(apply(binned_data, 1, function(x) zoo::rollsum(x, 5)));
```

### Topic model fit 

#### K=2

```{r echo=TRUE, eval=TRUE}
Topic_clus <- topics(sliding_binned_data, K=2)
K=2
barplot(t(Topic_clus$omega[order(elevation_metadata),]),col=2:(K+1),axisnames=F,space=0,border=NA,main="",las=1,ylim=c(0,1),
          cex.axis=1.5,cex.main=1.4)
  title(main=paste("Structure Plot,k=",K,"bins=",n_breaks))
  combo_patch_dir = paste0(elevation_metadata,"_",east_west_dir);
  combo_patch_dir_ordered = combo_patch_dir[order(elevation_metadata)];

  match_labs=match(unique(combo_patch_dir_ordered),combo_patch_dir_ordered);
  match_labs_suffix=c(match_labs[2:length(unique(combo_patch_dir_ordered))],35);
  match_labs_prefix=match_labs[1:(length(unique(combo_patch_dir_ordered)))];
  labs=match_labs_prefix + 0.5*(match_labs_suffix - match_labs_prefix);
  
  axis(1,at=labs,unique(combo_patch_dir_ordered),las=2);
  abline(v=match_labs[2:length(match_labs)]);
  

```

####  K=3

```{r echo=TRUE, eval=TRUE}
Topic_clus <- topics(sliding_binned_data, K=3)
K=3
east_west_elevation = paste0(metadata$WorE, "_", metadata$Elevation);

index1 <- which(metadata$WorE=="E");
index2 <- which(metadata$WorE=="W");
elevation1 <- metadata$Elevation[index1]; elevation2 <- metadata$Elevation[index2];
index_WE <- c(index1[order(elevation1)], index2[order(elevation2)]);

 K <- 3
    barplot(t(Topic_clus$omega[index_WE,]),col=2:(K+1),axisnames=F,space=0,border=NA,main=paste("No. of clusters=",3),las=1,ylim=c(0,1),cex.axis=1.5,cex.main=1.4)

    combo_patch_dir = paste0(east_west_elevation);
    combo_patch_dir_ordered = combo_patch_dir[index_WE];

  match_labs=match(unique(combo_patch_dir_ordered),combo_patch_dir_ordered);
  match_labs_suffix=c(match_labs[2:length(unique(combo_patch_dir_ordered))],35);
  match_labs_prefix=match_labs[1:(length(unique(combo_patch_dir_ordered)))];
  labs=match_labs_prefix + 0.5*(match_labs_suffix - match_labs_prefix);
  
  axis(1,at=labs,unique(combo_patch_dir_ordered),las=2);
  abline(v=match_labs[2:length(match_labs)]);
```
